\bf Estimating the number of clusters using cross-validation

### 2016-04-18 cvclust.tex
revise introduction; start revising theory

\documentclass[11pt]{article}
\usepackage{epsf}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath}
\usepackage{footnote}
\usepackage[bottom]{footmisc}
\usepackage[singlelinecheck=false]{caption}
\usepackage{caption}
\usepackage{amsthm}
\usepackage{enumitem}
\usepackage{graphics}
\usepackage{float}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{lscape}
\usepackage{array}
\textheight=8.7in
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{longtable}

\def\threedigits#1{%
  \ifnum#1<100 0\fi
  \ifnum#1<10 0\fi
  \number#1}

\topmargin=0.1in \oddsidemargin=-0.1cm \evensidemargin=-0.1cm

\paperheight=11in \paperwidth=8.5in \marginparwidth=0in

\marginparsep=0in \textwidth=6.5in \headheight=0in \headsep=0in

\onehalfspacing
\def\argmax{\mathop{\rm arg\,max}}
%\usepackage{xspace,epsfig,subfig}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{claim}{Claim}
\newtheorem{proposition}{Proposition}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newenvironment{sketch}{\noindent\emph{Proof Sketch:}}{$\quad \Box$}
%\newenvironment{proof}{\noindent\emph{Proof:}}{$\quad \Box$}

\bibpunct{(}{)}{;}{a}{,}{,}

% operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\PE}{\operatorname{PE}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\corr}{corr}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}
\newcommand{\T}{T}

@@

\begin{document}
\title{Estimating the number of clusters using Cross Validation}
\author{Wei Fu \qquad Patrick O. Perry \\\\ New York University}
\date{}
\maketitle
\begin{abstract}
Many clustering methods, including $k$-means, require the user to specify the
number of clusters as an input parameter. A variety of methods have been
devised to choose the number of clusters automatically, but they often rely on
strong modeling assumptions. We propose a data-driven approach to estimate the
number of clusters based on a novel form of cross-validation. This differs
from ordinary cross-validation, because clustering is fundamentally an
unsupervised learning problem. Simulation and real data analysis results show
that our proposed method outperforms existing methods, especially in
high-dimensional settings with heavy-tailed data.
\end{abstract}

\section{Introduction}

As a main task of exploratory data analysis, clustering organizes unlabeled
observations into groups such that observations in same group are more similar
compare to those in different group. Clustering is an important topic in
unsupervised learning because it can reveal the internal structure of data
through grouping, segment the data through partitioning and summarize data for
other purposes such as dimension reduction. It has being widely used in
various fields such as psychology, biology, statistics and machine learning
including pattern recognition, image segmentation etc.

After being proposed more than $50$ years, $k$-means remains one of the most
popular and widely used clustering algorithms \citep{jain2010data}. Like many
other clustering methods, $k$-means requires an input parameter $k$, the
number of clusters, to be specified by the user. Automatically and
quantitatively deciding such parameter is important and yet unsolved problem
\citep{fujita2014non}. Various methods have been proposed to tackle this
difficulty. One ad hoc approach is to explore the relationship between $W_k$
(within-cluster dispersion) and the number of cluster $k$ for a certain
clustering method such as $k$-means. Since $W_k$ decreases as $k$ increases,
one usually find the ``elbow" of curve obtain by plotting $W_k$ versus $k$ as
the appropriate number of clusters. The example on the top row of
Figure~\ref{fig:elbow} demonstrates such approach for data with $k=4$, where
the ``elbow" point indeed reveals the true number of clusters. This is based
on the idea that under partitioning data set has more impact than over
partitioning data set in terms of $W_k$. However, locating the ``elbow" point
is somewhat subjective and sometimes is not appropriate to select the optimal
$k$. The second example on the bottom row of Figure \ref{fig:elbow} shows a
situation where there is no clear choice of the ``elbow" point -- both $k=2$
and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can
never be selected as the optimal $k$ using such approach in this case since it
can hardly be viewed as the ``elbow" of the curve.

Recently, there are several new proposals to find the $k$ automatically. Gap
statistics \citep{tibshirani2001estimating} estimates $k$ by comparing the
change in within-cluster dispersion with that expected under an appropriate
reference null distribution. Specifically, the graph of $\log(W_k)$ is
compared with its expectation under an appropriate null reference distribution
of the data. The value of $k$ associated with the largest gap between
$\log(W_k)$ and the reference curve is selected as optimal $k$.
\citet{sugar2003finding} proposed an approach which finds the number of
clusters based on distortion, a quantity that measures the average distance,
per dimension. It's backed by a rigorous theoretical justification based on
information-theoretic ideas.   \citet{fraley2002model}'s Model-based method
employs the EM algorithm to estimate the parameters in Gaussian mixture model,
and select the best model ($k$) using BIC criterion. Stability-based criterion
is also proposed to locate the best $k$ by some authors such as
\citet{ben2001stability}, \citet{wang2010consistent} and
\citet{fang2012selection}. \citet{chiang2010intelligent} provides a nice review
of existing methods for finding the right $k$ in published literature.

Most existing methods are either model based method requires strong parameter
assumptions or otherwise lack of clear interpretation and theoretical justification.
Although many view selecting the number of clusters as a model selection problem, very few
approaches this problem from the prediction point of view. Select model with
smallest prediction error via cross-validation is one of the simplest and most
widely used model selection techniques in supervised learning. The lack of
true class (label) in data set makes the adoption of cross-validation into
unsupervised leaning problem difficult. One exception is
\cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction
strength. The strategy is to first cluster the test data and training data
into $k$ clusters respectively. Then, for each pair of observations that are
assigned to the same test cluster, algorithm determines whether they are also
assigned to the same cluster based on the training centers. The intuition here
is, if $k=k_0$, the true number of clusters, then the $k$ training set
clusters will be similar to the $k$ test clusters, and hence will predict them
well. However, a specifically defined prediction error measure is used in such
procedure, which is quite different from the one commonly used in
cross-validation procedure in supervised learning. Therefore the well
understood properties of cross-validation procedure in supervised learning
cannot be carried over to the prediction strength method, makes the
interpretation of its result difficult. \citet{wang2010consistent} also uses 
cross-validation to select the optimal $k$. Instead of selecting $k$ which minimizes 
prediction error, such method picks $k$ which minimizes the specifically defined clustering instability.

Our proposed method is a complete data-driven approach which doesn't rely on
any parametric assumptions. Through novel form of partitioning data set, we
are able to employing the cross-validation procedure in clustering exactly the
same way as in supervised learning problem. Hence, it's easy for reader to
understand the intuition behind our proposed method. Simulation and real data
application shows the superior performance of our proposed method compare with
existing methods in high-dimension settings and heavy-tailed data. Since the
embedded cross-validation procedure is well understood, it also makes our
method potentially easily to be extended in future study.

\section{Cross-validation for selecting the number of clusters}

Cross-validation is commonly used for model selection in supervised learning
problems.  In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
rows.  We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

@@

\subsection{Gabriel CV algorithm}
\label{sec:gabriel-cv-algorithm}

We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
cross-validation, we permute the rows and columns of the data matrix and then
partition the rows and columns as $N = n + m$ and $P = p + q$ for 
non-negative integers $n$, $m$, $p$, and $q$.  We treat the first $p$
columns as ``predictors'' and the last $q$ columns as ``responses'';
similarly, we treat the first $n$ rows as ``training'' and the last $m$ rows
as ``test''.  In block form, the permuted data matrix is
\[
  \dataX
  =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix},
\]
where
$\Xtrain \in \R^{n \times p}$,
$\Ytrain \in \R^{n \times q}$,
$\Xtest \in  \R^{m \times p}$,
and
$\Ytest \in  \R^{m \times q}$.

Given such a partition of $\dataX$, we perform four steps for each value of
$k$, the number of clusters:
\begin{enumerate}
  \item \label{step:gabriel-cluster}
    \textbf{Cluster:}
    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ and the
    cluster means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
    be the assigned cluster for row $i$.

  \item \label{step:gabriel-classify}
    \textbf{Classify:}
    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
    $\hGX : \R^p \to \{ 1, \dotsc, k \}$.

  \item \label{step:gabriel-predict}
    \textbf{Predict:}
    Apply the classifier to $X_{n+1}, \dotsc, X_{n+m}$, the rows of
    $\Xtest$, yielding predicted classes $\hGX_i = \hGX(X_i)$ for
    $i = n+1, \dotsc, n+m$.  For each value of $i$ in this range, compute
    predicted response $\hat Y_i = \bmuY(\hGX_i)$, where
    $\bmuY(g) = \bmuY_g$.

  \item \label{step:gabriel-evaluate}
    \textbf{Evaluate:}
    Compute the cross-validation error
    \[
      \CV(k) = \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2,
    \]
    where $Y_{n+1}, \dotsc, Y_{n+m}$ are the rows of $\Ytest$.
\end{enumerate}
\noindent
In principle, we could use any clustering and classification methods in
steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
report, we use $k$-means as the clustering algorithm.  For the classification
step, we compute the mean value of $X$ for each class; we assign an
observation to class $g$ if that class has the closest mean (randomly breaking
ties between classes).  The classification step is equivalent to linear
discriminant analysis with equal class priors and identity noise covariance
matrix.

@@

\section{Analysis of Gabriel Cross-Validation with Gaussian Noise}

\subsection{Single cluster in two dimensions}

Now we we analyze the asymptotic performance of Gabriel Cross-Validation, in
the case of Gaussian noise. Our main result is that with single-cluster
Gaussian data, if the predictor and response columns of $\dataX$ are weakly
correlated or independent, then the method will correctly prefer $k = 1$ to
$k = 2$ clusters. We state the result in the case where $\dataX$ has 2 columns,
but it is straightforward to generalize this result to higher dimensions.

\begin{proposition}\label{prop:1clust-2dim}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
independent draw from a mean-zero multivariate normal distribution with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
\end{proposition}

\begin{proof}

Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to $\{ Y_i
\}_{i=1}^{n}$. With $k = 1$, the single-cluster centroid will
be equal to $\bar Y_n$, the sample mean of the $Y_1, \dotsc, Y_n$, approximately equal to
$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
will be
\[
  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
         = 1 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
\]

Now we will consider the $k = 2$ case.  If $n$ is large enough, then
\citet{pollard1981strong} showed that the centroids $\bmuY_1$ and $\bmuY_2$
will be close to $\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) =
-\sqrt{2/\pi}$.  We have used Lemma~\ref{lem:truncated-normal-moments}
(Appendix~\ref{app:technical-lemmas}) to compute the expectations.  Further,
\citet{pollard1982central} showed that the errors will be of size $\OhP(n^{-1/2})$.

If $\rho > 0$ and $n$ is large enough, then classification rule learned from
$\{(X_i, \hGY_i)\}_{i=1}^{n}$ variables will be determined according to
whether $X > 0$; if $\rho < 0$ then the decision is according to whether $X <
0$.  More specifically, the decision boundary will be at $0 + \OhP(n^{-1/2})$.

In the $\rho > 0$ case, the cross-validation error will be
\begin{align*}
  \CV(2)
  &=
    \frac{1}{m}
    \sum_{i=n+1}^{n+m}
      \| (Y_i - \bmuY_1) 1\{\hGX_i = 1\} \|^2
      +
      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2
\\
  &=
    \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]
      + \OhP(m^{-1/2}) + \OhP(n^{-1/2}),
\end{align*}
where $a = \sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows
that $Y \mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$, so
that $\E[(Y - a)^2 \mid X] = (\rho X - a)^2 + (1 - \rho^2)$.  Applying
Lemma~\ref{lem:truncated-normal-moments}, we get that for large $m$ and $n$,
the Gabriel cross-validation error is close to
$1 + a^2 (1 - 2 \rho)$.

In the $\rho < 0$ case, a similar calculation shows that $\CV(2)$ is close to
$1 + a^2 (1 + 2 \rho)$.  In particular, if $|\rho| < 0.5$, then with
probability tending to $1$ and $m$ and $n$ increase, the asymptotic
cross-validation error for $k = 1$ will be smaller than for $k = 2$.
\end{proof}

@@

\begin{figure}[H]
\centering
\includegraphics[width=25pc]{demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

\subsection{Single cluster in higher dimensions}

To generalize the result of the previous section to higher dimensions, 
suppose that $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ is data from a single fold of
Gabriel Cross-Validation, with each $(X, Y)$ pair in $\R^p \times \R^q$ a
draw a mean-zero multivariate normal distribution. To analyze this setting, it
suffices to consider the mean-zero case with $\cov(X) = I_p$.

Suppose that $\cov(Y) = U \Lambda U^\T$ is the eigendecomposition of
$\cov(Y)$. For simplicity, suppose that the leading eigenvalue $\lambda_1$
has unit multiplicity, so that the first principal subspace of $Y$ is a line,
given by the span of the vector $u_1 \in \R^q$. When we apply $k$-means to
$Y_1, \dotsc, Y_n$, the centroids will be close to
$\bmuY_1 =  \sqrt{2\lambda_1 /\pi} u_1$ and
$\bmuY_2 = -\sqrt{2\lambda_1 /\pi} u_1$, with errors of size $\OhP(n^{-1/2})$.

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{float}
%\usepackage{footnote}
%\usepackage[bottom]{footmisc}
\usepackage[singlelinecheck=false]{caption}

\usepackage{graphicx,psfrag,epsf}
\usepackage[space]{grffile}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{-.3in}%
\addtolength{\topmargin}{-.8in}%

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{claim}{Claim}
\newtheorem{proposition}{Proposition}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}

% operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\PE}{\operatorname{PE}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}

@@

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if1\blind
{
  \title{\bf Estimating the number of clusters using cross-validation}
  \author{Wei Fu and Patrick O. Perry \\
  Stern School of Business, New York University}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Estimating the number of clusters using cross-validation}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Many clustering methods, including $k$-means, require the user to specify the
number of clusters as an input parameter. A variety of methods have been
devised to choose the number of clusters automatically, but they often rely on
strong modeling assumptions. We propose a data-driven approach to estimate the
number of clusters based on a novel form of cross-validation. This differs
from ordinary cross-validation, because clustering is fundamentally an
unsupervised learning problem. Simulation and real data analysis results show
that our proposed method outperforms existing methods, especially in
high-dimensional settings with heavy-tailed data.
\end{abstract}

\noindent%
{\it Keywords:} clustering, unsupervised learning, model selection
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

A clustering procedure segments a collection of items into smaller groups,
with the property that items in the same group are more similar to each other
than items in different groups \citep{hartigan1975clustering}.  These
procedures are used in two main applicaitons: (a) exploratory analysis, where clusters
reveal homogeneous sub-groups within a large sample \cite{TODO}; (b) data
reduction, where high-dimensional item attribute vectors get reduced discrete
cluster labels \citep{jain1999data}.

With many clustering methods, including the popular $k$-means clustering
procedure, the user must specify $k$, the number of clusters
\citep{jain2010data}.  One popular ad-hoc device for selecting the number of
clusters is to use an analogue of the principal components scree plot: plot
the within-cluster dispersion $W_k$, as a function of the number of clusters
$k$, looking for an ``elbow'' in the plot.  This approach is simple and often
performs well, but it requires subjective judgment as to where the elbow is
located, and as we demonstrate in Appendix~\ref{sec:elbow-fail}, the approach
can easily fail.  In this report, we propose a new method to choose the number
of clusters automatically.

The problem of choosing $k$ has been well-studied, and dozens of methods have
been proposed~\citep{chiang2010intelligent,fujita2014non}. The main difficulty
in choosing $k$ is that clustering is fundamentally an ``unsupervised''
learning problem, meaning that there is no obvious way to use ``prediction
ability'' to drive the model selection \citep{hastie2009elements}.  Most
existing methods for choosing $k$ instead rely on on explicit or implicit
assumptions about the data distribution, including it shape, scale, and
correlation structure.

Several authors advocate choosing $k$ by performing a sequence of hypothesis
tests will null and alternative hypotheses of the form $H_0 : k = k_0$ and
$H_1: k > k_0$, starting with $k_0 = 1$ and proceeding sequentially with
higher values of $k_0$ until a test fails to reject $H_0$. The gap statistic
method typifies this class of methods, with a test statistic that measures the
within-cluster dispersion relative to what is expected under a reference
distribution \citep{tibshirani2001estimating}. Similar methods include those
developed by \citet{TODO}.

Other authors have proposed choosing $k$ by using information
criteria.  \citet{sugar2003finding} proposed an approach that
minimizes the estimated ``distortion'', the average distance per dimension.
\citet{fraley2002model}'s model-based method fits Gaussian mixture model
models to the data, then selects the number of mixture components, $k$, using
the Bayesian Information Criterion (BIC).

A third set of approaches is based on the idea of ``stability'', that clusters
are meaningful if they manifest in multiple independent samples from the same
population. \citet{ben2001stability}, \citet{tibshirani2005cluster},
\citet{wang2010consistent} and \citet{fang2012selection} developed methods
based on this idea.

The procedure we propose in this report is based on a form of
cross-validation, and it is adaptive to the characteristics of the data
distribution. The essential idea is to devise a way to measure a form 
of internal prediction error associated each choose of $k$, and then choose
the $k$ with the smallest associated error.

\textit{TODO: Add outline}

\section{Cross-validation for selecting the number of clusters}
\label{sec:meth}
Cross-validation is commonly used for model selection in supervised learning
problems.  In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
rows.  We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

@@

\subsection{Gabriel CV algorithm}
\label{sec:gabriel-cv-algorithm}
We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
cross-validation, we permute the rows and columns of the data matrix and then
partition the rows and columns as $N = n + m$ and $P = p + q$ for 
non-negative integers $n$, $m$, $p$, and $q$.  We treat the first $p$
columns as ``predictors'' and the last $q$ columns as ``responses'';
similarly, we treat the first $n$ rows as ``training'' and the last $m$ rows
as ``test''.  In block form, the permuted data matrix is
\[
  \dataX
  =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix},
\]
where
$\Xtrain \in \R^{n \times p}$,
$\Ytrain \in \R^{n \times q}$,
$\Xtest \in  \R^{m \times p}$,
and
$\Ytest \in  \R^{m \times q}$.

Given such a partition of $\dataX$, we perform four steps for each value of
$k$, the number of clusters:
\begin{enumerate}
  \item \label{step:gabriel-cluster}
    \textbf{Cluster:}
    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ and the
    cluster means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
    be the assigned cluster for row $i$.
  \item \label{step:gabriel-classify}
    \textbf{Classify:}
    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
    $\hGX : \R^p \to \{ 1, \dotsc, k \}$.
  \item \label{step:gabriel-predict}
    \textbf{Predict:}
    Apply the classifier to $X_{n+1}, \dotsc, X_{n+m}$, the rows of
    $\Xtest$, yielding predicted classes $\hGX_i = \hGX(X_i)$ for
    $i = n+1, \dotsc, n+m$.  For each value of $i$ in this range, compute
    predicted response $\hat Y_i = \bmuY(\hGX_i)$, where
    $\bmuY(g) = \bmuY_g$.
  \item \label{step:gabriel-evaluate}
    \textbf{Evaluate:}
    Compute the cross-validation error
    \[
      \CV(k) = \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2,
    \]
    where $Y_{n+1}, \dotsc, Y_{n+m}$ are the rows of $\Ytest$.
\end{enumerate}
\noindent
In principle, we could use any clustering and classification methods in
steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
report, we use $k$-means as the clustering algorithm.  For the classification
step, we compute the mean value of $X$ for each class; we assign an
observation to class $g$ if that class has the closest mean (randomly breaking
ties between classes).  The classification step is equivalent to linear
discriminant analysis with equal class priors and identity noise covariance
matrix.

@@

\section{Analysis of Gabriel cross-validation with Gaussian noise}

\subsection{Single Cluster in two dimensions}

Now we we analyze the asymptotic performance of Gabriel Cross-Validation, in
the case of Gaussian noise. Our main result is that with single-cluster
Gaussian data in two-dimensions, if the predictor and response columns of $\dataX$ are weakly
correlated or independent, then the method will correctly prefer $k = 1$
clusters. In section~\ref{} we will generalize this result to 
cases where $\dataX$ has more than two columns.

\begin{theorem}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
independent draw from a mean-zero multivariate normal distribution with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
then $\CV(1) < \CV(k)$, $k > 1$ with probability tending to one as $m$ and $n$ increase.
\end{theorem}

\begin{proof}
Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, let $A_k = \{a_1,a_2,...,a_k\}$
denotes the set of cluster centers from applying $k$-means to $\{
Y_i\}_{i=1}^{n}$ with parameter $k$. Because $\{ Y_i\}_{i=1}^{n}$ is
symmetric, it's easy to see that the cluster centers are symmetric around $0$,
i.e. if $a^* \in A_k$ then $-a^* \in A_k$. The $k$ clusters are symmetric in
the same sense. Also, if $k$ is odd, then only one cluster contains both
negative and positive $Y$ and it's symmetric with center $0$. The rest have
pure positive or negative intervals; if $k$ is even, then all clusters have
pure positive or negative intervals.

Let $a_i$ denotes a cluster center with interval $[b_i, c_i]$. By above
argument, either $0 \leq b_i < c_i$ ($b_i < c_i \leq 0$) or $b_i = -c_i$ which
corresponding to $a_i=0$ with $k$ be odd. By symmetry, it's sufficient to only
consider case $0 \leq b_i < c_i$ (with $a_i > 0 $) and $b_i = -c_i$ (with $a_i
= 0 $) with $\rho \geq 0$.

From joint normality of $X$ and $Y$, it follows that $X \mid Y$ is normal with
mean $\rho Y$. Therefore, the mean value $\bar{X}_i$ of $X$ in cluster with
$Y$ center $a_i$ is

\begin{align*}
	\bar{X}_i &= E[X \mid b_i \leq Y \leq c_i] \\
			&= E[ E(X \mid Y) \mid b_i \leq Y \leq c_i] \\
			&= E[\rho Y \mid b_i \leq Y \leq c_i] \\
			&= \rho E[ Y \mid b_i \leq Y \leq c_i] \\
			&= \rho a_i
\end{align*}

That is, the mean values of $X$ in the $k$ clusters is $\{\rho a_1,\rho
a_2,...,\rho a_k\}$. Because we assign each observation to the closest
$\bar{X}_i$, $i=1,2,...,k$, it's easy to see the corresponding interval on $X$
for $\bar{X}_i = \rho a_i$ is $[\frac{\rho \left( a_{i-1}+a_i\right)}{2},
\frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$, where $a_{i-1}$ and $a_{i+1}$
are the two adjacent centers to $a_i$ on $Y$. Note that $[\frac{ \left(
a_{i-1}+a_i\right)}{2}, \frac{\left( a_{i}+a_{i+1}\right)}{2}] = [b_i, c_i]$
by the algorithm definition of $k$-means. The $Y$ center for such interval on
$X$ is

\begin{align*}
	\hat{\mu}^Y_i &= E[Y \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
	 &= E[ E(Y \mid X) \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
	 &= E[ \rho X \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
	 & =  \rho E[X \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
	  & =  \rho E[X \mid \rho b_i \leq X \leq \rho c_i] \\
	  &\leq \rho a_i
\end{align*}

where the equality holds if $b_i = -c_i$ or $\rho = 1$ on the last step. This is because

\begin{align*}
	 E[X \mid \rho b_i \leq X \leq \rho c_i] &= E[Y \mid \rho b_i \leq Y \leq \rho c_i] \\
	\mathtt{0 \leq b_i < c_i}, \hspace{0.05in} \mathtt{0 \leq \rho \leq 1} \hspace{0.1in} & \leq E[Y \mid b_i \leq Y \leq c_i] \\
	&= a_i
\end{align*}

The first equation above is because the same marginal distribution of $X$ and $Y$.

Note that Gabriel CV method predicts $Y$ to be $a_i$ for $X \in [\frac{\rho
\left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$,
and if $k=1$ the predicted $Y = 0$ for all $X$. Hence, to see if $CV(k) >
CV(1)$ for  $X \in [\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho
\left( a_{i}+a_{i+1}\right)}{2}]$, one only need to see if $ a_i-\hat{\mu}^Y_i
> \hat{\mu}^Y_i - 0$. Since $ \hat{\mu}^Y_i \leq \rho a_i$, it's clear that
$CV(k) > CV(1)$ whenever $\rho < 0.5$. This is true to all segment
$[\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left(
a_{i}+a_{i+1}\right)}{2}] = [\rho b_i, \rho c_i]$ except the one with $b_i =
-c_i$, in which $CV(k) = CV(1)$. Since $k>1$ and at most one segment has
$CV(k) = CV(1)$ while the rest have $CV(k) > CV(1)$ if $\rho < 0.5$, it is
clear that the overall prediction error $CV(k) > CV(1)$. By symmetry, we can
see $CV(k) > CV(1)$ if $|\rho| < 0.5$. The proof is complete.
\end{proof}

@@

\begin{figure}[H]
\centering
\includegraphics[width=3in]{demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

The reason why Gabriel CV method tends to select larger $k$ when correlation is high 
between dimensions is that it resembles the naive method mentioned in section \ref{sec:intro} 
under such circumstance. In fact, if $X$, $Y$ are perfectly correlated, Gabriel method is 
equivalent to the naive method.

\subsection{Single Cluster in higher dimensions}

\textit{todo: add and revise old proof}

\subsection{Two clusters in two dimensions}
